---
title: "Data transformation"
author: "Matteo Sostero"
output: github_document
editor_options: 
  chunk_output_type: console
---


## Packages for this session

|    package    |             purpose              |        installation        |
| ------------- | -------------------------------- | -------------------------- |
| `tidyverse`   | everything                       | CRAN                       |
| `nycflight13` | example datasets                 | CRAN                       |
| `lubridate`   | working with dates               | installed with `tidyverse` |

We need `tidyverse` and related dependencies.
The `nycflights13` package contains different example datasets, including *flights*


If not yet installed, download packages with `install.packages("name")`.
If asked, **do not compile from source**.

`CTRL/CMD + ENTER` to run the block of code below:

```{r eval=FALSE, message=FALSE, include=FALSE}
install.packages("tidyverse")
install.packages("nycflights13")
```

`CTRL/CMD + L` clears the console.

Load the packages for the session:
```{r}
# `tidyverse` loads the core packages of the *tidyverse* and shows their version.
library(tidyverse)
library(nycflights13)
library(lubridate) # tidyverse, not loaded by default
```


## Load example data

load the `flights` dataset from `nycflights13`:
```{r}
data(flights)
```

## Preview, `glimpse` and `View` data

Preview the data by invoking it: notice that not all variables are displayed for lack of space.
```{r}
flights
```

`glimpse` shows all variables as rows, variable types, and a preview of first few observations.
```{r}
glimpse(flights)
```

`View` shows the whole dataset in a dedicated tab of RStudio
```{r, eval=FALSE}
View(flights)
```


## `filter` rows matching conditions

Keep only observations such that month = 1 AND day = 1
```{r}
flights %>% filter(month == 1, day == 1)
```

Logical operators include:

* `==` strict equality
* `>`, `<`, `>=`, `<=` inequalities
* `%in%` to test if element belongs to list
* `is.na()` to test for missing value `NA`

Multiple conditions: 

* `!` as logical NOT
* `,` or `&` as logical AND

```{r}
flights %>% filter(month %in% c(2, 4))
```


## `select` and `pull` variables

We can select columns by name; the result is always a tibble.
Pressing TAB with the cursorinside `select( )` provides variable name completion. 
```{r}
flights %>% select(year, month, day, dep_time, dep_delay)
```

We can also select columns based on patterns in their names:
```{r}
flights %>% select(contains("dep"), starts_with("sched"))
```

or select intervals based on position (careful!)
```{r}
flights %>% select(year:day)
flights %>% select(1:4)
```

`pull()` returns a single column as a vector or values 
```{r, R.options=list(max.print=20)}
flights %>% pull(carrier)
```

We can combine this with `sort` (lexicographic sorting) and `unique` (deduplication) to see unique observations:
```{r}
flights %>% pull(carrier) %>% sort() %>% unique()
```


## `summarise` values

We can compute one or more variable summaries for the whole dataset (column-wise) using `summarise`.

`mean(., na.rm = TRUE)` computes the average value, excluding missing values.
```{r}
flights %>%
  summarise(
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE)
  )
```

How many missing values are there?
```{r}
flights %>%
  summarise(
    n_miss_dep_delay = sum(is.na(dep_delay)),
    n_miss_arr_delay = sum(is.na(arr_delay))
  )
```

This works because `is.na(dep_delay)` returns a *logical vector* (TRUE, FALSE) for all values of `dep_delay`. `sum` adds the TRUE values: 

```{r}
flights %>% pull(dep_delay) %>% is.na() %>% sum()
```


`group_by` summarises the values by groups:
```{r}
flights %>%
  group_by(carrier) %>%
  summarise(
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE)
  )
```


## `mutate` observations

Mutate adds new variables based on functions provided while preserving existing ones, 

```{r}
flights %>%
  mutate(
    distance_k = distance / 1000,
    air_time_h = air_time / 60
  ) %>%
  select(starts_with("distance"), starts_with("air_time"))
```


## Count observations with `n` and `count`

`n()` returns the total number of observation (in a group, if any).

```{r}
flights %>% group_by(carrier) %>% summarise(total_obs = n())
```

`tally` is short-hand for `summarise(n())`
```{r}
flights %>% group_by(carrier) %>% tally()
```

and `count` is short-hand for `group_by(carrier) %>% tally() %>% ungroup()`
```{r}
flights %>% count(carrier)
flights %>% count(carrier, dest)
```

Nevertheless, `n()` with `sum` or `length` can be used for composing functions inside `mutate` or `summarise`:

what share of flights arrived late, for each airline?
```{r}
flights %>% group_by(carrier) %>% summarise(sh_late_arr = sum(arr_delay > 0, na.rm = TRUE) / n())
```


## `arrange` to sort rows

`arrange` sorts observations by lexicographic order of one or more variables. `desc` arranges in descending order

```{r}
flights %>% count(carrier) %>% arrange(desc(n))
```

`top_n` sorts and returns the top n entries by value
```{r}
flights %>% top_n(10, arr_delay)
```


## Keep unique rows with `distinct`

`distinct` finds unique combinations of  *rows in tibbles*, while `unique` finds unique *elements in vectors*

What are the destination from each airport?
```{r}
flights %>% distinct(origin, dest)
```

Keep *all columns* with unique combinations of (origin, dest). 
Notice that this keeps the **first** observation of duplicated columns:
```{r}
flights %>% distinct(origin, dest, .keep_all = TRUE)
```


## `join` (merge) tibbles by value

`airlines` in `nycflights13` contains the full names of airlines, reported as `carrier`in flights
```{r}
data("airlines")
airlines
```

Do `airlines` and `flights` have variables in common
```{r}
names(flights)
names(airlines)

intersect(names(flights), names(airlines))
```

What variables are in `airlines` but not in `flights` (and viceversa)?
```{r}
setdiff(names(airlines), names(flights))
setdiff(names(flights), names(airlines))
```

They have `carrier` is in common. Does it have common values between tibbles?
```{r}
airlines %>% pull(carrier) %>% unique() %>% sort()
flights %>% pull(carrier) %>% unique() %>% sort()

intersect(
  airlines %>% pull(carrier) %>% unique(),
  flights %>% pull(carrier) %>% unique()
)
```

Then we can join the tibbles, to add `name` (full carrier name) from `airlines` to `flights`.

`left_join(x, y, by)` return all rows from x where there are matching values in y, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned.

See also `right_join`, `full_join`, `inner_join`, `anti_join`...
we use `select(new_name = name)` to rename variables on the spot.

```{r}
left_join(flights, airlines, by = "carrier") %>%
  select(call_sign = carrier, carrier_name = name)
```

Did we miss anything? Compare results with `anti_join`

```{r}
anti_join(airlines, flights, by = "carrier")
```

Let's add a ficticious "SA" "Sant'Anna" airline and see:
```{r}
airlines_sa <- airlines %>%
  add_row(carrier = "SA", name = "Sant'Anna Airlines")

anti_join(airlines_sa, flights, by = "carrier")
```
