---
title: "Data transformation with dplyr"
author: "Matteo Sostero"
output: github_document
editor_options: 
  chunk_output_type: console
---
  
## Load the tidyverse package

```{r echo=TRUE, message=FALSE}
library(tidyverse)
```

## Load example data
The *nycflights13* package contains different example datasets, including *flights*

```{r}
library(nycflights13)
data(flights)
```

## Preview, `glimpse` and `View` data

Preview the data by invoking it: notice that not all variables are displayed for lack of space.
```{r}
flights
```

`glimpse` shows all variables as rows, variable types, and a preview of first few observations.
```{r}
glimpse(flights)
```

`View` shows the whole dataset in a dedicated tab of RStudio
```{r}
View(flights)
```


## `filter` rows matching conditions

Keep only observations such that month = 1 AND day = 1
```{r}
flights %>% filter(month == 1, day == 1)
```

Logical operators include:

* `==` strict equality
* `>`, `<`, `>=`, `<=` inequalities
* `%in%` to test if element belongs to list
* `is.na()` to test for missing value `NA`

Multiple conditions: 

* `!` as logical NOT
* `,` or `&` as logical AND

```{r}
flights %>% filter(month %in% c(1,2))
```


# `select` and `pull` variables

We can select columns by name; the result is always a tibble.
Pressing TAB with the cursorinside `select( )` provides variable name completion. 
```{r}
flights %>% select(year, month, day, dep_time)
```

We can also select columns based on patterns in their names:
```{r}
flights %>% select(contains("dep"), starts_with("sched"))
```

`pull()` returns a single column as a vector or values 
```{r}
flights %>% pull(carrier)
```

We can combine this with `sort` (lexicographic sorting) and `unique` (deduplication) to see unique observations:
```{r}
flights %>% pull(carrier) %>% sort() %>% unique()
```


## `summarise` values

We can compute one or more variable summaries for the whole dataset (column-wise) using `summarise`.

`mean(., na.rm = TRUE)` computes the average value, excluding missing values.

```{r}
flights %>% 
  summarise(
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE)
  )
```

How many missing values are there?
```{r}
flights %>% 
  summarise(
    n_miss_dep_delay = sum(is.na(dep_delay)),
    n_miss_arr_delay = sum(is.na(arr_delay))
    )
```

This works because `is.na(dep_delay)` returns a *logical vector* (TRUE, FALSE) for all values of `dep_delay`. `sum` adds the TRUE values: 

```{r}
flights %>% pull(dep_delay) %>% is.na() %>% sum()
```


`group_by` summarises the values by groups:
```{r}
flights %>% 
  group_by(carrier) %>% 
  summarise(
    n_miss_dep_delay = sum(is.na(dep_delay)),
    n_miss_arr_delay = sum(is.na(arr_delay))
  )
```


## `mutate` observations

Mutate adds new variables based on functions provided while preserving existing ones, 

```{r}
flights %>% 
  mutate(
    distance_k = distance / 1000,
    air_time_h = air_time / 60
  ) %>% 
  select(starts_with("distance"), starts_with("air_time"))
```

# `arrange` to sort rows

`arrange` sorts observations by lexicographic order of one or more variables. `desc` arranges in descending order

```{r}
flights %>% arrange(desc(carrier), dep_time)
```

`top_n` sorts and returns the top n entries by value

```{r}
flights %>% top_n(3, arr_delay)
```


## `join` (merge) tibbles by value

`airlines` in `nycflights13` contains the full names of airlines, reported as `carrier`in flights
```{r}
data("airlines")
airlines
```

`left_join(x, y, by)` return all rows from x where there are matching values in y, and all columns from x and y. If there are multiple matches between x and y, all combination of the matches are returned.

See also `right_join`, `full_join`, `inner_join`, `anti_join`...

```{r}
left_join(flights, airlines, by = "carrier") %>% 
  select(call_sign = carrier, carrier_name = name)
```







