---
title: "Data types, lists, and mapping"
author: "Matteo Sostero"
output: github_document
editor_options: 
  chunk_output_type: console
---


## Packages for this session

|    package    |             purpose              |        installation        |
| ------------- | -------------------------------- | -------------------------- |
| `tidyverse`   | everything                       | CRAN                       |
| `nycflight13` | example datasets                 | CRAN                       |
| `lubridate`   | working with dates               | installed with `tidyverse` |
| `glue`        | pasting and interpreting strings | installed with `tidyverse` |


If not yet installed, download packages with `install.packages("name")`.
If asked, **do not compile from source**.
```{r eval=FALSE, message=FALSE, include=FALSE}
install.packages("tidyverse")
install.packages("nycflights13")
```

Load the packages for the session:
```{r}
# `tidyverse` loads the core packages of the *tidyverse* and shows their version.
library(tidyverse)
library(nycflights13)
library(lubridate) # tidyverse, not loaded by default
library(glue) # tidyverse, not loaded by default
```

## Variable types
`glimpse` shows all variable names and `<types>`, including:

* `<int>` **integer**: *signed* (positive or negative) integer
* `<dbl>` **double**: "double-precision" real number
* `<chr>` **character**: character string
* `<fct>` **factor**: categorical variable
* `<dttm>` **date-time**

```{r}
flights %>% glimpse()
```

Inspect variable types with typeof:
```{r}
flights %>% pull(dep_time) %>% typeof()
flights %>% pull(dep_delay) %>% typeof()
flights %>% pull(carrier) %>% typeof()
flights %>% pull(time_hour) %>% typeof()
```

What are object types are there in R?

```{r}
# Numbers are doubles by default…
typeof(1.0)
typeof(1)

# …unless specified as integers with `L` or coerced
typeof(1L)
as.integer(1) %>% typeof()

# strings:
typeof("a")
```


## Vectors: *atomic vector* and *list* classes

The main type of *R* objects are *vectors*, either:

1. **lists**: (heterogeneous) contain elements of different type;
2. **atomic vectors**: contain elements of same type:
    1. *numeric*: *integer* or *double*
    2. *character*
    3. *logical*
    4. *complex*
    5. *raw*
    
We can construct **atomic vectors** by *concatenating* `,`-sepatated elements with `c( )` 

Atomic vectors:
```{r}
# numeric double
typeof(c(1, 2, 3))

# numeric integer
typeof(c(1L, 2L, 3L))

# …character…
typeof(c("a", "b"))
typeof(c("1", "2"))

# …logical…
typeof(c(TRUE, FALSE))
typeof(NA)
typeof(c(1, 2) == 1)
```

We can initialise empty **lists** of lenght *n* with `vector("list", n)` and assign elements by positions
```{r}
ex_list <- vector("list", 4) # initialise list

ex_list # empty list

# assign elements to list by position:
ex_list[[1]] <- 10L # a lonely integer in the 1st slot
ex_list[[2]] <- c("a", "b") # an atomic character vector in 2nd slot
ex_list[[3]] <- c(1.1, 2.2) # an atomic double vector in 3rd slot
ex_list[[4]] <- c(NA, TRUE, FALSE) # an atomic logical vector in 3rd slot

ex_list # populated list
```


## Applying functions to vector (list) elements with `map`

`map(.x, .f)` applies a function `.f` to every element of a list or atomic vector `.x`
Functions can be:

* already defined in R: `sum`, `mean`, `filter`, `mutate`, ecc.
* defined in place as “anonymous functions” `.f` with the syntax `~ .`, where `~` introduces a function and `.` is the argument
```{r}
# a list of three elements: 2, 4, and 6
list(2, 4, 6)
# every list element is actually a lenght-1 atomic vector

# we map the list elements .x to a function .f = ~ .^2
# the function is introduced by ~ and takes the argument . and raises ^ to the power of 2
map(.x = list(2, 4, 6), .f = ~ .^2)

# equivalent to piping the list as first argument of map(.)
# and omit the argument names .x and .f
list(2, 4, 6) %>% map(.f = ~ .^2)
```

we can map lists containing atomic vectors of different lenghts:
```{r}
# A list of three elements, containing numeric vectors of different lenghts
list(c(1, 2), c(3, 4, 5, 6), c(7, 8, 9))

# we map functions that allow vector arguments, and returns a single element
list(c(1, 2), c(3, 4, 5, 6), c(7, 8, 9)) %>% map(~ sum(.))
list(c(1, 2), c(3, 4, 5, 6), c(7, 8, 9)) %>% map(~ mean(.))
```

we can map functions to list of string vectors as well.
For instance, `collapse` takes a vector of character and pastes them together
```{r}
list(c("a", "b", "c"), c("hello", "bye"))
list(c("a", "b", "c"), c("hello", "bye")) %>% map(~ collapse(.))
```

`tibble` and `data.frame` objects are also `list`! Hence, we can `map` them:

```{r, R.options=list(max.print=20)}
is.list(flights) # the tibble is also a list
length(flights) # a list of 19 elements (the variables, or columns)

# let's take the first elements of the “list” flights:
flights[[1]]
# it's a numeric vector, with as many elements as observations (rows) of flights
```
 
We can `map` the variables that make up `flights`
```{r}
# what is the type of each variable
flights %>% map(~ typeof(.))

# is each variable of type integer?
flights %>% map(~ is.integer(.))

# find number of elements in each variable
flights %>% map(~ length(.))

# count missing elements in each variable
flights %>% map(~ sum(is.na(.)))
```


## Factors

`factor` variables are convenient to work with categorical values. 
We can transform (typically `character`) objects with `factor`

```{r}
flights %>%
  pull(carrier) %>%
  head(n = 10) # show only first 10 elements

carriers <- flights %>% pull(carrier) %>% factor()

levels(carriers)
flights %>% pull(carrier) %>% sort() %>% unique()

typeof(carriers) # factors are stored as type integer
```


We can manually set factor levels:
```{r}
c("a", "b") %>% factor(levels = c("a", "b")) # levels all observed in the vector
c("a", "b") %>% factor(levels = c("a", "b", "c")) # as before, and one more level
c("a", "b", "z") %>% factor(levels = c("a", "b", "c")) # levels in data are different
```

Assign meaningful labels to factor levels. By default, levels are *unique, sorted* values, unless we provide a custom order
```{r}
c("a", "b") %>% factor() # by default, levels and labels are unique sorted values
c("a", "b") %>% factor(levels = c("b", "a")) # custom level order
```

We can change how levels are labelled by setting `labels` (in the same order as `levels`):
```{r}
c("a", "b") %>% factor(labels = c("alpha", "beta")) # "a" = "alpha" and "b" = "beta"
c("a", "b") %>% factor(labels = c("beta", "alpha")) # mind the order! "a" = "beta" and "b" = "alpha"

# better carefully set both levels and labels:
c("a", "b") %>% factor(levels = c("b", "a"), labels = c("beta", "alpha"))
```


Example: work within `flights` tibble to change `carrier` to factor:
```{r}
flights %>%
  mutate(carrier = factor(carrier)) %>%
  select(year:day, carrier)
```

since airlines uniquely maps carrier codes in airline names, we can use that to set factor levels
```{r}
flights %>%
  mutate(
    carrier = factor(
      carrier,
      levels = airlines %>% pull(carrier),
      labels = airlines %>% pull(name)
    )
  ) %>%
  select(year:day, carrier)
```

in shorhand, using `$` to select a variable in a tibble
```{r}
flights %>%
  mutate(carrier = factor(carrier, levels = airlines$carrier, labels = airlines$name)) %>%
  select(year:day, carrier)
```


## Date and time variables

Dates are generally difficult! `lubridate` package (in `tidyverse`, but not loaded automatically) makes it easier:

The functions `ymd()` `dmy()` etc., allow to parse strings containing "Year, Month, Date" values (in a given order)

```{r}
library(lubridate) # installed with tidyverse, but not loaded by default

ymd("2018-05-17") # parse date Y-M-D

dmy("15/05/2018") # input date in different order…
mdy("05.15.2018") # …and with different separators

mdy("05.15.2018") == dmy("15/05/2018") # they are parsed to identical representations

mdy("May 17th 2018") # Can parse natural language
dmy("17 Maggio 2018") # also in non-English other locales!

# Parse heterogeneous formats (still Y-M-D), but written differently
c(
  20090101, "2009-01-02", "2009 01 03", "2009-1-4",
  "2009-1, 5", "Created on 2009 1 6", "200901 !!! 07"
) %>% ymd()
```

Operations with dates:
```{r}
typeof(ymd("2018-05-17")) # actually a double number previewed in human-redable format

dmy("16/05/2018") > dmy("15/05/2018") # > means "after"

today() # current date

# extract components:
today() %>% day()
today() %>% month()
today() %>% quarter()

# also week day (as integer or factor in system locale)
today() %>% wday()
today() %>% wday(label = TRUE)

today() %>% leap_year() # is it a leap year?
today() %>% dst() # is it Daylight Savings Time?
```


### Date arithmetic

Date arithmetic is hard! Lots of conventions and unspoken assumptions. What does "a month from now" mean exactly?

```{r}
today() + days(1) # one day from now
today() + months(1) # one month from now

# Round dates:
today() %>% floor_date(unit = "month") # round down to first day of month
today() %>% ceiling_date(unit = "month") # round up to first day of next month
today() %>% ceiling_date(unit = "month") - days(1) # round up to last day of this month
today() %>% rollback() # round down to last day of previous month
today() %>% rollback(roll_to_first = T) # round down first day of same month
```


### Construct dates from separate columns:
```{r}
flights %>% select(year, month, day) # separate year, month, day (integer) variables

flights %>%
  select(year, month, day) %>%
  unite(date, year, month, day, sep = "-") # paste variables together as (dash-separated) character

flights %>%
  select(year, month, day) %>%
  unite(date, year, month, day, sep = "-") %>%
  mutate(date = ymd(date))
```


What's in a date? Obviously *day*, *month*, *year*. But also time zone! (by default UTC)
```{r}
dmy("16/05-2018", tz = "Europe/Rome") == dmy("16/05-2018", tz = "Europe/London")
```


### Higher- and lower- resolution dates and time

`dmy_hms` (and obvious permutations) parse date and times to seconds:
```{r}
# Using different representations:
dmy_hms("17/05/2018T8:22,30")
dmy_hms("17/05/2018T8:22.30")
dmy_hms("17-05-2018 8:22:30")
dmy_hms("17-05-2018 8-22-3")
dmy_hms("17-05-2018 8-22-3")
```

`hms` parses time-of-day values:
```{r}
# Using different representations:
hms("12:30:00")
hms("12.31.00")
```

Both have variants for lower-resolution input:
```{r}
# parse date and time to resolution of minutes
dmy_hm("17-05-2018 8-22")
# parse date and time to precision of hours
dmy_h("17-05-2018 08")

# time at diffrent resolution
hm("12.31") # with different separators
ms("2.05") # with different separators
```
