---
title: "Data input and output"
author: "Matteo Sostero"
output: github_document
editor_options: 
  chunk_output_type: console
---

**Packages needed for this session**:

* `tidyverse`
* `readxl` (installed with tidyverse)
* `haven` (installed with tidyverse)
* `glue` (installed with tidyverse)
* `naniar` **new!**

```{r, eval=FALSE, message=FALSE, include=FALSE}
install.packages("naniar")
```

## Load the packages for this session
`tidyverse` loads (eight) packages of the *tidyverse* and shows their version.
 Also load `readxl` to read Excel files and `haven` for importing and exporting Stata, SAS and SPSS data.

```{r}
library(tidyverse)
library(readxl)
library(haven)
library(naniar)
library(glue)
```


## Data input and output
Import a table with `read_csv` (NB, different than `read.csv`!). 
`read_csv` parses variables by guessing the column types based on heuristics:
```{r}
flights_import <- read_csv("./data/flights.csv")

glimpse(flights_import)
```

inspect the data using `miss_var_summary` from `naniar` to see how many are missing:
```{r}
flights_import %>% miss_var_summary()

# which variables have missing values?
#QUIZ

# plot of missing values by variables
flights_import %>% gg_miss_var()
```

## Solving parsing failures

Let's try with a (pathologically) tricky file
```{r}
challenge <- read_csv("./data/challenge.csv")
```

what's wrong? let's inspect:
```{r}
problems(challenge)
```

open the raw `csv` file to see what's happening (hint: line 1001):
```{r eval=FALSE, message=FALSE, include=FALSE}
rstudioapi::navigateToFile("data/challenge.csv", line = 1001)
```

we help `read_csv` parser by scanning more rows before guessing type:
```{r}
challenge <- read_csv("./data/challenge.csv", guess_max = 1500)
```

we can also specify column types manually:
```{r}
challenge <- read_csv("./data/challenge.csv", col_types = list(x = col_double(), y = col_date()))
```

_pro-tip__: import problematic columns as character and inspect them:
```{r, R.options=list(max.print=20)}
challenge <- read_csv("./data/challenge.csv", col_types = list(x = col_character(), y = col_character()))

challenge %>% pull(y) %>% sort() %>% unique()
challenge %>% pull(x) %>% sort() %>% unique()
```

`readr` tries to guess the type of object based on heuristics: 
```{r}
guess_parser("123")
guess_parser("1,234")
guess_parser(c(".", "-"))
guess_parser(c("10W", "20N"))
guess_parser("10:30")
guess_parser("15/06/2018")
```


## Import from internet

Read a file from a URL:
```{r}
flights_import <- read_csv("https://raw.githubusercontent.com/CoMoS-SA/workshop-R-tidyverse/master/data/flights.csv")
```


## Importing Excel files with `readxl`
We can import Excel workbooks with `read_excel`: it works with `.xlsx` and `.xls` files. 
```{r}
# Enumerate sheets in the workbook
excel_sheets("./data/nycflights.xlsx")

# Access a given sheet
flights_xl  <- read_excel("./data/nycflights.xlsx", sheet = "Flights")
airlines_xl <- read_excel("./data/nycflights.xlsx", sheet = "Airlines")
```
NB: the package `readxl` does not provide functions to *export* Excel workbooks, but the package `openxls` does.


## Importing and exporting Stata `.dta` files using `haven`
```{r}
# read .dta file. 
# Default text encoding in Stata >= 14 is utf-8, but before also be "windows-1252"

mtcars_stata <- read_dta("./data/mtcars.dta")
mtcars_stata <- read_dta("./data/mtcars.dta", encoding = "UTF-8")

write_dta(mtcars_stata, "./data/mtcars.dta", version = 13)
```


## Saving data in R
The best format for saving R objects for later use is `.rds`

```{r}
write_rds(mtcars_stata, "./data/mtcars.rds")
mtcars_stata <- read_rds("./data/mtcars.rds")
```



## Applying functions to vector (list) elements with `map`

`map()` applies a function to every element of a vector (atomic or list).
Functions can be:

* already defined in R: `sum`, `mean`, `filter`, `mutate`, ecc.
* defined in place as “anonymous functions” with the syntax `~ .`, where `~` introduces a function and `.` is the argument

```{r}
list(1, 2, 3)

map(list(1, 2, 3), ~ .^2)

list(1, 2, 3) %>%  map(., ~ .^2) # anonymous function elevating square 
list(1, 2, 3) %>%  map(~ .^2)
```

`tibbles` and `data.frames` are `lists` too! Hence, we can `map` them

```{r}

```



## Reading and merging all files in a directory

Sometimes data is fractioned in disparate files, differing only by the value of a variable.
In the `./data/scattered/` directory, I fractioned the `flights` data by month.

Enumerate files in that directory:
```{r}
list.files("./data/scattered/") # bare file names
list.files("./data/scattered/", full.names = TRUE) # relative path and file names
```

Read files and assign them as elements of a list:
```{r}
flights_frac <- list.files("./data/scattered/", full.names = TRUE) %>% 
  map(~ read_csv(.))
```

Append all the list elements (tibbles) in a single tibble
```{r}
flights_frac <- list.files("./data/scattered/", full.names = TRUE) %>% 
  map(~ read_csv(.)) %>% 
  bind_rows()
```


`map_df` maps the function directly into a tibble
```{r}
flights_frac <- list.files("./data/scattered/", full.names = TRUE) %>% 
  map_df(~ read_csv(.))
```


