---
title: "Data input and output"
author: "Matteo Sostero"
output: github_document
editor_options: 
  chunk_output_type: console
---

## Load the packages for this session
`tidyverse` loads (eight) packages of the *tidyverse* and shows their version.

```{r}
library(tidyverse)
```

## Data input and output

Import a table with `read_csv` (NB, different than `read.csv`!). 

`read_csv` parses variables by guessing the column types based on heuristics:

```{r}
flights_import <- read_csv("./data/flights.csv")

glimpse(flights_import)
```

Let's try with a tricky file
```{r}
challenge <- read_csv("./data/challenge.csv")
```

what's wrong? let's inspect:
```{r}
problems(challenge)
```

we help `read_csv` parser by scanning more rows before guessing type:
```{r}
challenge <- read_csv("./data/challenge.csv", guess_max = 1500)
```

we can also specify column types manually:
```{r}
challenge <- read_csv("./data/challenge.csv", col_types = list(x = col_double(), y = col_date()))
```

_pro-tip__: import problematic columns as character and inspect them:
```{r}
challenge <- read_csv("./data/challenge.csv", col_types = list(x = col_double(), y = col_character()))

challenge %>% pull(y) %>% sort() %>% unique()
```

`readr` tries to guess the type of object based on heuristics: 
```{r}
guess_parser("123")
guess_parser("1,234")
guess_parser(c(".", "-"))
guess_parser(c("10W", "20N"))
guess_parser("10:30")
```




